---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# assessEpidemicCurves

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/jpmeagher/assessEpidemicCurves.svg?branch=master)](https://travis-ci.com/jpmeagher/assessEpidemicCurves)
<!-- badges: end -->

The goal of assessEpidemicCurves is to model epidemic curves under heterogeneous disease reproduction, providing estimates for the time-varying reproduction number and assessing epidemic curves for evidence of superspreading. This package accompanies [Reference]

## Installation

You can install the development version of assessEpidemicCurves from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jpmeagher/assessEpidemicCurves")
```

## Case Study: COVID-19 in the Republic of Ireland

```{r}
library(assessEpidemicCurves)
library(ggplot2)
```


We consider the COVID-19 epidemic in the Republic of Ireland from March 1, 2020 to 28 February, 2021.

```{r, echo = FALSE, fig.height=3.5, fig.width=7}
ggplot(covid_incidence_roi_epidemiological_date) +
  geom_bar(
    aes(x = date, y =  count), stat = "identity",
    alpha = 0.5
  ) +
  theme_classic() +
  labs(
    title = "COVID-19 in the Republic of Ireland",
    x = "Epidimiological Date",
    y = "Reported Cases",
    color = NULL
  )
```

Analyses are based on a hierarchical model for daily incidence count of COVID-19 \(\boldsymbol y = \left(y_0, y_1, \dots, y_N \right)^\top\)
